Carbon Nucleosynthesis

### 2023-04-29 thesis/tex/main.tex
fine adjustments

\documentclass[12pt,oneside]{report}

% \AddToHook{cmd/section/before}{\clearpage}

@@

\usepackage{aas_macros}
\usepackage[letterpaper, top=1in]{geometry}
\usepackage{titlesec}

@@

\usepackage{isotope}
% \usepackage[titles]{tocloft} % fig titles
% \renewcommand{\cftfigpresnum}{\figurename\enspace}
% \setlength{\cftfignumwidth}{5em}
% \renewcommand{\cfttabpresnum}{\tablename\enspace}
% \setlength{\cfttabnumwidth}{5em}

@@

\begin{document}

%\doublespacing

@@

% Abstract of the paper
\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}
% context
Stellar evolution models provide uncertain predictions for elemental yields due to uncertainties in stellar physics.
% aims
Here, I constrain the nucleosynthetic yields of C using multizone Galactic chemical evolution models.
% results
By matching the slopes of the \caah\ and \caafe\ relationships in a sample of APOGEE subgiant stars, I find that CCSNe and AGB stars contribute \about{80\%} and \about{20\%} of total C production respectively.
% 
To also match the normalization of the trends, I estimate the CCSNe C/Mg yield ratio $\Ycc/\Yoc = 1.57 + 0.59 \left(Z/Z_\odot\right)$, which explains the \caah\ trend when including the AGB C contribution.
%
Variations in the star formation history only slightly impact the low [Mg/Fe] tail of the [C/Mg]-[Mg/Fe] relation. However, most of the scatter in the \caah\ distribution can be attributed to measurement errors and radial migration. 
% 
Due to the degeneracy between the normalization of elemental yields and the strength of mass-loading in Galactic outflows, I constrain only the {\it relative} yield of C and Mg from a simple stellar population and the metallicity dependence thereof. While measurements of gas-phase C abundances are challenging, my model is broadly consistent with the [C/O]-[O/H] trend observed in compiled literature measurements.

@@

Galactic chemical evolution (GCE) aims to understand the chemical and star formation histories of galaxies. At the core of GCE models are stellar yields---the amount of each chemical element stars produce (or destroy). Each element is produced in different amounts by different stars, leaving traces of the Galaxy's evolutionary history. Through \textit{Galactic Archeology}, we can reconstruct the history of our galaxy by investigating clues left behind in stars.

Here, I aim to understand C enrichment---where it is produced and how its abundances evolve over time. C is unique nucleosynthetically, one of few light elements (along with N) produced in asymptotic giant branch (AGB) stars \citep[e.g.][]{jennifer19, KL14}. C and N are well-studied elements since they are easy to observe in stellar spectra, even at the lowest metallicities \cite[e.g.][]{fabbian+09, nissen+14, lambert81, laird85, lambert86}. Additionally, C and N are used as age indicators in red giant branch stars \citep{martig16, MG15, hasselquist19, vincenzo+21}.

Gas-phase [C/O] ratios---observed in very low metallicity, high redshift damped Lyman-alpha systems (DLAs)---decrease with increasing [O/H] \citep{FN15, cooke+17}. Then, [C/O] increases above $\rm [O/H] \approx -1$ (\citealt{berg+19}; see discussion in \S\ref{sec:gas}).
While we know C is produced in AGB stars and core-collapse supernovae (CCSNe), we still have a limited understanding of the magnitude and metallicity-dependence of each process.

Despite their central role in GCE models, nucleosynthetic yields predicted by stellar evolution models are rife with uncertainties. The production rates of elements in stars are shaped by poorly understood processes, including mass loss, opacity, nuclear reaction rates, rotational mixing, and convection \citep{KL14,ventura+13, LC18}.
Changes in massive stars' explodability also introduce further variation and uncertainty in population-averaged yields \citep{emily+21}.

GCE models typically use unmodified nucleosynthetic predictions and attempt to create models matching observations by varying the star formation history (SFH) and other evolutionary parameters. Here, I introduce C yields as an additional free parameter, determining which yield prescriptions reproduce Galactic abundance trends.
\cite{james+23} examined similar GCE models of N, an element whose production is closely related to C. \cite{james+23} found that trends in N and O abundances can be explained by the metallicity dependence of relative N and O yields. Here, I extend their models to C and show that similar constraints on C and O relative yields can be obtained from observed abundance ratios. Additionally, I assess which yield prescriptions reproduce Galactic abundance trends while assessing the impact of GCE model assumptions, such as SFH and outflow mass loading.

Accurate determination of stellar birth abundances for C and N is challenging. When a star enters the red giant branch (RGB), material from the CNO-processed core is mixed with the envelope in first dredge up (FDU), enhancing N and depleting C \citep{iben67, vincenzo+21,KL14}. Measurements of these evolved stars'  atmospheres will no longer reflect their birth abundances.  Additionally, gas-phase measurements of C are extremely limited as C lacks strong lines in HII regions \citep{skillman+20}.

To this end, I use a sample of subgiant stars from APOGEE \citep{apogee17}. According to stellar evolution theory and observations \citep{gilroy89, korn+07, lind+08, souto+18, souto19}, these stars have not yet experienced FDU but have well-mixed photospheres. Their photosphoric abundances should thus represent their birth composition.  In Fig.~\ref{fig:subgiants}, I plot this sample, selected by the criteria in Roberts et al. (2023, in prep; see Appendix \ref{sec:jack}). These empirical results contain information on the relative C and Mg yields, which I use later in this work.

\begin{figure}[htp]
    \includegraphics{subgiants_mgh.pdf}
    \includegraphics{subgiants_mgfe.pdf}
    \caption[APOGEE Subgiants]{The [C/Mg] ratio against [Mg/H] (top) and [Mg/Fe] (bottom) for the \citetjack~sample of APOGEE subgiants. On the top, I plot high and low-$\alpha$ stars in blue and orange, using the separation defined in Equation \ref{eq:high_alpha}. On the bottom, I color-code stars according to their [Mg/H] abundance.}
    \label{fig:subgiants}
\end{figure}

@@

Theoretical models of stellar nucleosynthesis provide the starting point of this investigation. I focus on three primary nucleosynthetic pathways: asymptotic giant branch (AGB) stars, core-collapse supernovae (CCSNe), and type Ia supernovae (SNe Ia). Each process has unique timescales and yields, traceable through the tools of GCE. I compare C, produced in CCSNe and AGB stars, to Mg and Fe as tracers of CCSNe and SNe Ia respectively.

When a single stellar population (SSP) forms, CCSNe are the first enrichment event. CCSNe  explode within $\lesssim 40$ Myr, providing light elements (C, O, and Mg) and heavier elements (Fe and beyond). Metallicity-independent yields from CCSNe are the only statistically significant source of O and Mg ($\alpha$ elements). Next, low-mass stars begin to reach the end of their lives. By shedding their outer layers, AGB stars are important sources of C, N, and neutron capture elements.  Finally, white dwarfs explode, releasing Fe and other iron-peak elements.

For an element X and star with mass $M$, the stellar yield $\tilde{y}$ is defined as the net production of X relative to $M$, or
\begin{equation}
    \tilde{y}_{\rm X} = \frac{M_{\rm X,\ ejected} - Z_{0, X} M_{\rm ejected}}{M}   
\end{equation}
where $M_{\rm ejected}$ and $M_{\rm X, ejected}$  are the total ejected mass of the envelope and the element X, respectively. For example, if a 1 $M_\odot$ star has $\tilde{y}_{\rm C} = 10^{-3}$, then the star will add $10^{-3}\ M_\odot$ of C to the interstellar medium (ISM). 
Although per-star yields are necessary to compute AGB star enrichment rates in GCE models, IMF-averaged\footnote{IMF: Initial Mass Function} yields are useful in interpreting their predictions. For a yield $y$ from a star of mass M and initial metallicity $Z$, the IMF-averaged yield is given by 
\begin{equation} \label{eq:imf-yield}
    y_{\rm X}(Z) = 
    \int_{M_{\rm min}}^{M_{\rm max}} 
    \tilde{y}_{\rm X}(M, Z)
    \frac{dN}{dM}  \ dM
\end{equation}
where ${dN}/{dM}$ is the normalized IMF, and $M_{\rm min}$ and $M_{\rm max}$ are the minimum and maximum mass of stars, which I take to be $0.08 M_{\sun}$ and $100 M_{\sun}$ respectively. For AGB stars, the yields are truncated above 8 $M_{\odot}$. 
To calculate the IMF-weighted net yields, I use the Versatile Integrator for Chemical Evolution code (\VICE\footnote{\VICE~is available at \url{https://github.com/giganano/VICE}}).

To focus on C yields, I adapt the yield choices of other elements from \citet{james+21, james+23}.
Table \ref{tab:fiducial_mod} contains my fiducial yields, in units of a SSP's birth mass.
Also following \citet{james+21, james+23}, I take the SNeIa delay time distribution to be a
$t^{-1.1}$ power-law suggested by the observations of \citet{maoz+12}.

\begin{table}
	\centering
    \caption[Fiducial Model]{Yields for the fiducial model in units of SSP birh mass. See \S\ref{sec:agb} for the definition of C11.}
	\label{tab:fiducial_mod}

@@

An AGB star is a low mass ($\lesssim 8 M_{\sun}$) star during its final phase of evolution.  In an AGB star, two competing processes determine the outcome of C production: third dredge up (TDU) and hot bottom burning (HBB).  TDU accompanies thermal pulses in AGB stars. During a TDU event, material from the CO core is mixed with material exposed to partial He-shell burning, increasing C and O abundances in the envelope \citep{KL14}. If this envelope is lost during the AGB phase, then C yields are enhanced.
HBB is the activation of proton capture reactions, and the CNO cycle%
\footnote{The CNO (I) cycle is
\isotope[12]{C}(p, $\gamma$)
\isotope[13]{N}($\beta^+ \nu_e$)
\isotope[13]{C}(p, $\gamma$)
\isotope[14]{N}(p, $\gamma$)
\isotope[15]{O}($\beta^{+}\nu_e$)
\isotope[15]{N}(p, \alp)
\isotope[12]{C}.
 Other minor branches of the CNO cycle are of minor importance to C yields
 \citep{solar-fusion}.
}
at the base of the convective envelope. Because the $^{14}$N(p, $\gamma$) proton capture is the slowest component of the CNO cycle \citep{solar-fusion}, the CNO cycle converts nearly all \isotope[12]{C} into \isotope[14]{N}.
As a result, when both TDU and HBB occur, \isotope[12]{C} yields are lowered (see discussion in \citealt{james+23} and \citealt{ventura+13}).

In this work, I explore four different sets of AGB star yield tables from literature, providing well-sampled grids in metallicity and mass for use in chemical evolution models. 
\begin{itemize}
    \item C11: \citet{cristallo+11, cristallo+15}
    \item K10: \citet{karakas10}
    \item V13: \citet{ventura+13,ventura+14,ventura+18, ventura+20}
    \item K16: \citet{KL16} \citet{karakas+18}
\end{itemize}
I use C11 table, amplified by a factor of 2.9, as the fiducial AGB yield.

Fig.~\ref{fig:y_agb} compares the AGB C yield $\tilde{y}_\text{C}^\text{AGB}(M, Z)$ for each model I consider. 
The yields may be negative if a star returns material to the ISM with a lower C abundance than the material the star was formed from.
Most models agree on the qualitative shape of the net fractional AGB C yield---% 
stars with masses between \about{2~M_\odot} and \about{4~M_\odot} have the highest fractional C yields, with the mass of the peak increasing and overall yields decreasing with increasing $Z$.  High mass, high $Z$ stars destroy \isotope[12]{C} because they experience both TDU and HBB, but HBB is much more efficient.

In Fig.~\ref{fig:agb-ssp}, I plot the total production of C by AGB stars in a SSP at an age $t$. 
Effectively, this is equivalent to integrating Eq.~\ref{eq:imf-yield} down to $M_{to}$, the turnoff mass for time $t$.
({In our model, the mass-lifetime relation is
$\log \tau_M = 1.02 - 3.57\log M + 0.90 \left(\log M\right)^2$,
where $\tau_M$ is in Gyr,
from \citealt{larson74}.})
As the mass range $2\lesssim M \lesssim 4$ is most important for C production, about half of C production occurs before \about{1} Gyr, similar to SNe Ia Fe. 
K10 and K16 weight C production more heavily towards massive stars resulting in a faster enrichment delay time, whereas the C11 and V13 models predict a slightly longer timescale of \about{1} Gyr, but little to no C is produced more than 2 Gyr after a star formation event. This is in contrast to Fe whos production lasts up to 10 Gyr after a star formation event. As shown in the right panel of Fig.~\ref{fig:agb-ssp}, with increasing $Z$, C enrichment occurs earlier, and C destruction in low-mass stars leads to a declining C abundance at late times.

\begin{figure}
    \centering
 	    \includegraphics[scale=1]{agb_yields.pdf}
        \caption[AGB C yields]{The net fractional AGB C yield  plotted as a function of initial stellar mass $M$ and color-coded according to metallicity. The black dashed line shows $\tilde{y}=0$ for reference. Each panel represents yields from one of four AGB models: C11, K10, V13, K16 (see \S \ref{sec:agb}). }
        \label{fig:y_agb}
\end{figure}
\begin{figure}
    \includegraphics[scale=1]{y_agb_t2.pdf}

    \caption[AGB yields delay time distribution]{
        C production by AGB stars as a function of SSP age, normalized to the total mass $M_{\rm tot}$ produced at $t=10$ Gyr. Left: The four AGB yield models from literature at solar metallicity (C11, K10, V13, or K16). The delay time distribution of type Ia supernovae ($\propto t^{-1.1}$) is plotted as a dashed black line for comparison. Right: The C11 AGB model at different metallicities. }
    \label{fig:agb-ssp}
\end{figure}

Fig.~\ref{fig:yagb-z} shows IMF-averaged C yields for each AGB model as a function of metallicity $Z$.
V13 differs in that it shows a non-monotonic metallicity dependence. However, this effect is only for models with $\log Z/Z_\odot \lesssim -1$.
Otherwise, models differ only in their yield normalization and metallicity dependence. All models predict yields within a factor of \about{2} for fixed metallicity.
For example, the three models C11, K10, and K16 predict $y_\text{C}^\text{AGB}$ to be between 0.006 and 0.008 at solar metallicity, but C11 has a much shallower metallicity dependence than the K10 and K16 models. V13 instead predicts a yield \about{0.004}.

@@

    \caption[AGB yield metallicity dependence]{The net fractional IMF-weighted AGB C yield $\Ycagb$ as a function of metallicity for each of our AGB yield models.
    }
    \label{fig:yagb-z}
\end{figure}

@@

Massive stars form $^{12}$C in their cores through the triple--$\alpha$ process. However, only C ejected through supernovae and stellar winds contributes to the yield. 
While there are many stellar models providing predictions of CCSNe yields, the results of these models are highly uncertain due to the many stellar modeling uncertainties.

In Fig.~\ref{fig:y_cc}, I plot calculations of the IMF-integrated yields, defined with Eq.~\ref{eq:imf-yield} (computed using \VICE's \texttt{vice.yields.ccsne.fractional} function). 
CCSNe models predict a wide range of C yields, spanning almost 1 dex. 
Both the \citet{NKT13} and \cite{LC18} models show positive metallicity dependence. 
The \cite{LC18} models also include rotation, showing that variations in the rotational velocity of the star can dramatically increase C production and the metallicity dependence of $\Ycc$. Rotation induces more mixing allowing the CO core to grow larger. Our fiducial model's steep metallicity dependence near $Z\approx Z_\odot$ could be explained by rotation. 
Fig.~\ref{fig:y_cc} shows the C11 model for comparison on the top. Especially at $Z\approx Z_\odot$, most CCSNe models dominate AGB C production. I will later also show empirically that CCSNe should dominate C production.

On the bottom of Fig.~\ref{fig:y_cc}, I also show the CCSNe-[C/Mg] ratio, defined by
\begin{equation}\label{eq:c_mg_cc}
    {\rm [C/Mg]^{CC}} = \log_{10}\left( \frac{\Ycc}{\Yoc}\right) - \log_{10} \left( \frac{Z_{{\rm C},\ \sun }}{Z_{{\rm Mg},\ \sun }} \right).
\end{equation}
${\rm [C/Mg]^{CC}}$ describes what [C/Mg] would be if CCSNe were the only process producing C.
Once again, different CCSNe models span a large range in [C/Mg]. 
I chose to instead parameterize $\Ycc$ to enable agreement with observations, as most CCSNe models fail to achieve near-solar [C/Mg].
Assumptions about the explodability landscape affect C and Mg production. Increasing the fraction of stars that explode increases $\Ycc$, as stars that directly collapse do not contribute to explosive yields \citep{emily+21}. However, C is relatively unaffected by the black-hole landscape, as very massive stars contribute carbon through enriched winds. Since Mg is formed deeper in the core of massive stars, the Mg yield drops much more steeply, so models, where few stars explode (S16/W18), result in much higher [C/Mg].

CCSNe models also do not reach [O/Mg] due to overproduction of O, or underproduction of Mg, or both \citep{emily+21}. Here, I assume ${\rm [O/Mg]} = 0$, which is not compatible with CCSNe models but is consistent with APOGEE observations \citep{weinberg+19, weinberg+22}.

>>>>

%\DocumentMetadata{testphase=phase-III,testphase=math}
\documentclass[12pt,oneside,letterpaper]{report}

@@

\usepackage{aas_macros}
%\usepackage[letterpaper, top=1in]{geometry}
\usepackage{titlesec}
\usepackage{microtype}

@@

\usepackage{isotope}

\usepackage{accsupp}
\usepackage[accsupp]{axessibility}
\pdfcompresslevel=0

% accessibility
% \usepackage{tagpdf}
% \usepackage[tagpdf]{axessibility}

\usepackage[]{acro}
\usepackage{array}

\acsetup{
    make-links = true,
    list/template = tabular,
    templates/colspec = {p{0.2\linewidth} p{0.8\linewidth} },
    pages/display = none,
    format/short = {\sc}, 
}

\DeclareAcronym{cc}{
    short = ccsn,
    long = core collapse supernova,
    short-plural = e,
    long-plural = e,
    extra = {massive star explosions}
}
\DeclareAcronym{ia}{
    short = sne ia,
    long = supernovae type Ia,
    extra = {exploding white dwarfs}
}

\DeclareAcronym{agb}{
    short = agb,
    long = asymptotic giant branch,
    short-plural = s,
    long-plural = \ stars,
}
\DeclareAcronym{gce}{
    short = gce,
    long = galactic chemical evolution
}

\DeclareAcronym{dla}{
    short = {dla},
    long = {damped Lyman-alpha},
    short-plural = s,
    long-plural = \ systems,
}

\DeclareAcronym{apogee}{
    short = {apogee},
    long = {Apache Point Observatory Galactic Evolution Experiment},
    extra = {a large optical spectroscopic survey of Milky Way stars}
}

\DeclareAcronym{sfh}{
    short = sfh,
    long = star formation history,
}
\DeclareAcronym{ssp}{
    short = ssp,
    long = single stellar population,
}
\DeclareAcronym{imf}{
    short = imf,
    long = initial mass function,
}

\newcommand{\cc}{\ac{cc}}
\newcommand{\agb}{\ac{agb}}
\newcommand{\ia}{\ac{ia}}

\newcommand{\sfh}{\ac{sfh}}
\newcommand{\ssp}{\ac{ssp}}
\newcommand{\imf}{\ac{imf}}
\newcommand{\gce}{\ac{gce}}
\newcommand{\apogee}{\ac{apogee}}

@@

\begin{document}
%\doublespacing

@@

% Abstract of the paper
\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}
% context
Stellar evolution models provide uncertain predictions for elemental yields due to uncertainties in stellar physics.
% aims
Here, I constrain the nucleosynthetic yields of C using multizone Galactic chemical evolution models.
% results
By matching the slopes of the \caah\ and \caafe\ relationships in a sample of APOGEE subgiant stars, I find that massive stars and low mass stars contribute \about{80\%} and \about{20\%} of total C production respectively.
% 
To also match the normalization of the trends, I estimate the massive star C/Mg yield ratio $\Ycc/\Yoc = 1.57 + 0.59 \left(Z/Z_\odot\right)$, which explains the \caah\ trend when including the low mass C contribution.
%
Variations in the star formation history only slightly impact the low [Mg/Fe] tail of the [C/Mg]-[Mg/Fe] relation. However, most of the scatter in the \caah\ distribution can be attributed to measurement errors and radial migration. 
% 
Due to the degeneracy between the normalization of elemental yields and the strength of mass-loading in Galactic outflows, I constrain only the {\it relative} yield of C and Mg from a simple stellar population and the metallicity dependence thereof. While measurements of gas-phase C abundances are challenging, my model is broadly consistent with the [C/O]-[O/H] trend observed in compiled literature measurements.

@@

\Ac{gce} aims to understand the chemical and star formation histories of galaxies. At the core of \ac{gce} models are stellar yields---the amount of each chemical element stars produce (or destroy). Each element is produced in different amounts by different stars, leaving traces of the Galaxy's evolutionary history. Through \textit{Galactic Archeology}, we can reconstruct the history of our galaxy by investigating clues left behind in stars.

Here, I aim to understand C enrichment---where it is produced and how its abundances evolve over time. C is unique nucleosynthetically, one of few light elements (along with N) produced in \ac{agb} stars \citep[e.g.][]{jennifer19, KL14}. C and N are well-studied elements since they are easy to observe in stellar spectra, even at the lowest metallicities \cite[e.g.][]{fabbian+09, nissen+14, lambert81, laird85, lambert86}. Additionally, C and N are used as age indicators in red giant branch stars \citep{martig16, MG15, hasselquist19, vincenzo+21}.

Gas-phase [C/O] ratios---observed in very low metallicity, high redshift \acp{dla}---decrease with increasing [O/H] \citep{FN15, cooke+17}. Then, [C/O] increases above $\rm [O/H] \approx -1$ (\citealt{berg+19}; see discussion in \S\ref{sec:gas}).
While we know C is produced in \ac{agb} stars and \acp{cc}, we still have a limited understanding of the magnitude and metallicity-dependence of each process.

Despite their central role in \gce{} models, nucleosynthetic yields predicted by stellar evolution models are rife with uncertainties. The production rates of elements in stars are shaped by poorly understood processes, including mass loss, opacity, nuclear reaction rates, rotational mixing, and convection \citep{KL14,ventura+13, LC18}.
Changes in massive stars' explodability also introduce further variation and uncertainty in population-averaged yields \citep{emily+21}.

\gce{} models typically use unmodified nucleosynthetic predictions and attempt to create models matching observations by varying the \sfh{} and other evolutionary parameters. Here, I introduce C yields as an additional free parameter, determining which yield prescriptions reproduce Galactic abundance trends.
\cite{james+23} examined similar \gce{} models of N, an element whose production is closely related to C. \cite{james+23} found that trends in N and O abundances can be explained by the metallicity dependence of relative N and O yields. Here, I extend their models to C and show that similar constraints on C and O relative yields can be obtained from observed abundance ratios. Additionally, I assess which yield prescriptions reproduce Galactic abundance trends while assessing the impact of \gce{} model assumptions, such as \sfh{} and outflow mass loading.

Accurate determination of stellar birth abundances for C and N is challenging. When a star enters the red giant branch, material from the CNO-processed core is mixed with the envelope in first dredge up, enhancing N and depleting C \citep{iben67, vincenzo+21,KL14}. Measurements of these evolved stars'  atmospheres will no longer reflect their birth abundances.  Additionally, gas-phase measurements of C are extremely limited as C lacks strong lines in HII regions \citep{skillman+20}.

To this end, I use a sample of subgiant stars from the \apogee{} \citep{apogee17}. According to stellar evolution theory and observations \citep{gilroy89, korn+07, lind+08, souto+18, souto19}, these stars have not yet experienced first dredge up but have well-mixed photospheres. Their photosphoric abundances should thus represent their birth composition.  In Fig.~\ref{fig:subgiants}, I plot this sample, selected by the criteria in Roberts et al. (2023, in prep; see Appendix \ref{sec:jack}). These empirical results contain information on the relative C and Mg yields, which I use later in this work.

\begin{figure}[htp]
    \includegraphics{subgiants_mgh.pdf}
    \includegraphics{subgiants_mgfe.pdf}
    \caption[APOGEE Subgiants]{The [C/Mg] ratio against [Mg/H] (top) and [Mg/Fe] (bottom) for the \citetjack~sample of \apogee{} subgiants. On the top, I plot high and low-$\alpha$ stars in blue and orange, using the separation defined in Equation \ref{eq:high_alpha}. On the bottom, I color-code stars according to their [Mg/H] abundance.}
    \label{fig:subgiants}
\end{figure}

@@

Theoretical models of stellar nucleosynthesis provide the starting point of this investigation. I focus on three primary nucleosynthetic pathways: \acf{agb} stars, \acf{cc}, and type Ia supernovae \acf{ia}. Each process has unique timescales and yields, traceable through the tools of \gce. I compare C, produced in \cc{} and \ac{agb} stars, to Mg and Fe as tracers of \cc{} and \acf{ia} espectively.

When a \ssp~forms, \cc{} are the first enrichment event. \cc{}  explode within $\lesssim 40$ Myr, providing light elements (C, O, and Mg) and heavier elements (Fe and beyond). Metallicity-independent yields from \cc{} are the only statistically significant source of O and Mg ($\alpha$ elements). Next, low-mass stars begin to reach the end of their lives. By shedding their outer layers, \ac{agb} stars are important sources of C, N, and neutron capture elements.  Finally, white dwarfs explode, releasing Fe and other iron-peak elements.

For an element X and star with mass $M$, the stellar yield $\tilde{y}$ is defined as the net production of X relative to $M$, or
\begin{equation}
    \tilde{y}_{\rm X} = \frac{M_{\rm X,\ ejected} - Z_{0, X} M_{\rm ejected}}{M}   
\end{equation}
where $M_{\rm ejected}$ and $M_{\rm X, ejected}$  are the total ejected mass of the envelope and the element X, respectively. For example, if a 1 $M_\odot$ star has $\tilde{y}_{\rm C} = 10^{-3}$, then the star will add $10^{-3}\ M_\odot$ of C to the interstellar medium. 
Although per-star yields are necessary to compute \ac{agb} star enrichment rates in \gce{}  models, \imf-averaged\footnote{IMF: Initial Mass Function} yields are useful in interpreting their predictions. For a yield $y$ from a star of mass M and initial metallicity $Z$, the \imf-averaged yield is given by 
\begin{equation} \label{eq:imf-yield}
    y_{\rm X}(Z) = 
    \int_{M_{\rm min}}^{M_{\rm max}} 
    \tilde{y}_{\rm X}(M, Z)
    \frac{dN}{dM}  \ dM
\end{equation}
where ${dN}/{dM}$ is the normalized \imf, and $M_{\rm min}$ and $M_{\rm max}$ are the minimum and maximum mass of stars, which I take to be $0.08 M_{\sun}$ and $100 M_{\sun}$ respectively. For \ac{agb} stars, the yields are truncated above 8 $M_{\odot}$. 
To calculate the \imf-weighted net yields, I use the Versatile Integrator for Chemical Evolution code (\VICE\footnote{\VICE~is available at \url{https://github.com/giganano/VICE}}).

To focus on C yields, I adapt the yield choices of other elements from \citet{james+21, james+23}.
Table \ref{tab:fiducial_mod} contains my fiducial yields, in units of a \ssp's birth mass.
Also following \citet{james+21, james+23}, I take the \ac{ia} delay time distribution to be a
$t^{-1.1}$ power-law suggested by the observations of \citet{maoz+12}.

\begin{table}
	\centering
    \caption[Fiducial Model]{Yields for the fiducial model in units of \ssp~birth mass. See \S\ref{sec:agb} for the definition of C11.}
	\label{tab:fiducial_mod}

@@

An \ac{agb} star is a low mass ($\lesssim 8 M_{\sun}$) star during its final phase of evolution.  In an \ac{agb} star, two competing processes determine the outcome of C production: third dredge up and hot bottom burning.  Third dredge up accompanies thermal pulses in \ac{agb} stars. During a third dredge up event, material from the CO core is mixed with material exposed to partial He-shell burning, increasing C and O abundances in the envelope \citep{KL14}. If this envelope is lost during the \ac{agb} phase, then C yields are enhanced.
Hot bottom burning is the activation of proton capture reactions, and the CNO cycle%
\footnote{The CNO (I) cycle is
\isotope[12]{C}(p, $\gamma$)
\isotope[13]{N}($\beta^+ \nu_e$)
\isotope[13]{C}(p, $\gamma$)
\isotope[14]{N}(p, $\gamma$)
\isotope[15]{O}($\beta^{+}\nu_e$)
\isotope[15]{N}(p, \alp)
\isotope[12]{C}.
 Other minor branches of the CNO cycle are of minor importance to C yields
 \citep{solar-fusion}.
}
at the base of the convective envelope. Because the $^{14}$N(p, $\gamma$) proton capture is the slowest component of the CNO cycle \citep{solar-fusion}, the CNO cycle converts nearly all \isotope[12]{C} into \isotope[14]{N}.
As a result, when both third dredge up and Hot bottom burning occur, \isotope[12]{C} yields are lowered (see discussion in \citealt{james+23} and \citealt{ventura+13}).

In this work, I explore four different sets of \ac{agb} star yield tables from literature, providing well-sampled grids in metallicity and mass for use in chemical evolution models. 
\begin{itemize}
    \item C11: \citet{cristallo+11, cristallo+15}
    \item K10: \citet{karakas10}
    \item V13: \citet{ventura+13,ventura+14,ventura+18, ventura+20}
    \item K16: \citet{KL16} \citet{karakas+18}
\end{itemize}
I use C11 table, amplified by a factor of 2.9, as the fiducial \ac{agb} yield.

Fig.~\ref{fig:y_agb} compares the \ac{agb} C yield $\tilde{y}_\text{C}^\text{AGB}(M, Z)$ for each model I consider. 
The yields may be negative if a star ejects material with a lower average C abundance than the material the star was formed from.
Most models agree on the qualitative shape of the net fractional \ac{agb} C yield---% 
stars with masses between \about{2~M_\odot} and \about{4~M_\odot} have the highest fractional C yields, with the mass of the peak increasing and overall yields decreasing with increasing $Z$.  High mass, high $Z$ stars destroy \isotope[12]{C} because they experience both third dredge up and hot bottom burning, but the latter is much more efficient.

In Fig.~\ref{fig:agb-ssp}, I plot the total production of C by \ac{agb} stars in a \ssp{} at an age $t$. 
Effectively, this is equivalent to integrating Eq.~\ref{eq:imf-yield} down to $M_{to}$, the turnoff mass for time $t$.
({In our model, the mass-lifetime relation is
$\log \tau_M = 1.02 - 3.57\log M + 0.90 \left(\log M\right)^2$,
where $\tau_M$ is in Gyr,
from \citealt{larson74}.})
As the mass range $2\lesssim M \lesssim 4$ is most important for C production, about half of C production occurs before \about{1} Gyr, similar to \ac{ia} Fe. 
K10 and K16 weight C production more heavily towards massive stars resulting in a faster enrichment delay time, whereas the C11 and V13 models predict a slightly longer timescale of \about{1} Gyr, but little to no C is produced more than 2 Gyr after a star formation event. This is in contrast to Fe whos production lasts up to 10 Gyr after a star formation event. As shown in the right panel of Fig.~\ref{fig:agb-ssp}, with increasing $Z$, C enrichment occurs earlier, and C destruction in low-mass stars leads to a declining C abundance at late times.

\begin{figure}
    \centering
 	    \includegraphics[scale=1]{agb_yields.pdf}
        \caption[Low mass C yields]{The net fractional \ac{agb} C yield  plotted as a function of initial stellar mass $M$ and color-coded according to metallicity. The black dashed line shows $\tilde{y}=0$ for reference. Each panel represents yields from one of four \ac{agb} models: C11, K10, V13, K16 (see \S \ref{sec:agb}). }
        \label{fig:y_agb}
\end{figure}
\begin{figure}
    \includegraphics[scale=1]{y_agb_t2.pdf}

    \caption[C yields delay time distribution]{
    C production by \ac{agb} stars as a function of \ssp{} age, normalized to the total mass $M_{\rm tot}$ produced at $t=10$ Gyr. Left: The four \ac{agb} yield models from literature at solar metallicity (C11, K10, V13, or K16). The delay time distribution of type Ia supernovae ($\propto t^{-1.1}$) is plotted as a dashed black line for comparison. Right: The C11 \ac{agb} model at different metallicities. }
    \label{fig:agb-ssp}
\end{figure}

Fig.~\ref{fig:yagb-z} shows \imf-averaged C yields for each \ac{agb} model as a function of metallicity $Z$.
V13 differs in that it shows a non-monotonic metallicity dependence. However, this effect is only for models with $\log Z/Z_\odot \lesssim -1$.
Otherwise, models differ only in their yield normalization and metallicity dependence. All models predict yields within a factor of \about{2} for fixed metallicity.
For example, the three models C11, K10, and K16 predict $y_\text{C}^\text{AGB}$ to be between 0.006 and 0.008 at solar metallicity, but C11 has a much shallower metallicity dependence than the K10 and K16 models. V13 instead predicts a yield \about{0.004}.

@@

    \caption[Low mass yield metallicity dependence]{The net fractional \imf-weighted \ac{agb} C yield $\Ycagb$ as a function of metallicity for each of our \ac{agb} yield models.
    }
    \label{fig:yagb-z}
\end{figure}

@@

Massive stars form $^{12}$C in their cores through the triple--$\alpha$ process. However, only C ejected through supernovae and stellar winds contributes to the yield. 
While there are many stellar models providing predictions of \cc{} yields, the results of these models are highly uncertain due to the many stellar modeling uncertainties.

In Fig.~\ref{fig:y_cc}, I plot calculations of the \imf-integrated yields, defined with Eq.~\ref{eq:imf-yield} (computed using \VICE's \texttt{vice.yields.ccsne.fractional} function). 
\cc{} models predict a wide range of C yields, spanning almost 1 dex. 
Both the \citet{NKT13} and \cite{LC18} models show positive metallicity dependence. 
The \cite{LC18} models also include rotation, showing that variations in the rotational velocity of the star can dramatically increase C production and the metallicity dependence of $\Ycc$. Rotation induces more mixing allowing the CO core to grow larger. Our fiducial model's steep metallicity dependence near $Z\approx Z_\odot$ could be explained by rotation. 
Fig.~\ref{fig:y_cc} shows the C11 model for comparison on the top. Especially at $Z\approx Z_\odot$, most \cc{} models dominate \ac{agb} C production. I will later also show empirically that \cc{} should dominate C production.

On the bottom of Fig.~\ref{fig:y_cc}, I also show the \cc{}-[C/Mg] ratio, defined by
\begin{equation}\label{eq:c_mg_cc}
    {\rm [C/Mg]^{CC}} = \log_{10}\left( \frac{\Ycc}{\Yoc}\right) - \log_{10} \left( \frac{Z_{{\rm C},\ \sun }}{Z_{{\rm Mg},\ \sun }} \right).
\end{equation}
${\rm [C/Mg]^{CC}}$ describes what [C/Mg] would be if \cc{} were the only process producing C.
Once again, different \acp{cc} models span a large range in [C/Mg]. 
I chose to instead parameterize $\Ycc$ to enable agreement with observations, as most \acp{cc} models fail to achieve near-solar [C/Mg].
Assumptions about the explodability landscape affect C and Mg production. Increasing the fraction of stars that explode increases $\Ycc$, as stars that directly collapse do not contribute to explosive yields \citep{emily+21}. However, C is relatively unaffected by the black-hole landscape, as very massive stars contribute carbon through enriched winds. Since Mg is formed deeper in the core of massive stars, the Mg yield drops much more steeply, so models, where few stars explode (S16/W18), result in much higher [C/Mg].

\acp{cc} models also do not reach [O/Mg] due to overproduction of O, or underproduction of Mg, or both \citep{emily+21}. Here, I assume ${\rm [O/Mg]} = 0$, which is not compatible with \acp{cc} models but is consistent with \apogee{} observations \citep{weinberg+19, weinberg+22}.
